---
title: "Exploratory Analysis - Visuals"
author: "Evan Canfield"
date: "5/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose
The following notebooks is intended for exploratory analysis in the creation of a budget projection workflow.

# Import
## Packages
```{r import_packages}
# Ensure package installation
if (!require("pacman")) install.packages("pacman")

# Install packages
pacman::p_load(
  dplyr,
  DT,
  ggthemes,
  highcharter,
  janitor,
  lubridate,
  plotly,
  readxl
)

```

## Data
```{r import_data}
start_date <- "2021-01-01"
end_date <- "2022-06-01"
# end_date <- NA


projections_raw <- fund_projections(
  start_date = start_date,
  end_date = end_date
  ) 

projections_raw %>% head()
```


# Shiny App Objects
## Categories
```{r}
category_filter <- category_key %>% 
  pull(category) %>% 
  sort()

cat(category_filter)
```

## Output Table
### Table Formatting
```{r}
df_table <- projections_raw %>% 
  dplyr::filter(category == category_filter[17])

  df <- df_table %>%
  dplyr::ungroup() %>% 
    dplyr::select( year, month, available_funds) %>% 
    dplyr::rename(
      Year = year,
      Month = month,
      'Available Funds' = available_funds
    )

df
```

### DataTable Formatting
```{r}
 DT::datatable(
      data = df, 
      fillContainer = FALSE,
      options = list(
        columnDefs = list(
          list(
            className = 'dt-center',
            targets = "_all"
            )
          ),
        lengthMenu = c(20, 50, 100),
        pageLength = 20
      )
      ) %>%
      DT::formatCurrency(
        columns = 'Available Funds',
        currency = '$',
        mark = ",",
        digits = 0
      ) %>%
      DT::formatRound(
        columns = 'Year',
        digits = 0,
        mark = ""
      )
```


## Static Plot
### To Do
* Dashed lines on January of every year, if not first or last month
  * Will need a loop
* Fix cropping of final month
* Left Align x-axis labels

### Data
```{r}
df_static <- projections_raw %>% 
  dplyr::filter(category == category_filter[17])
```

### Plots
```{r}
p0 <- ggplot(
  data = df_static, 
  mapping = aes(
    x = date, 
    y = available_funds
    )
)

p1 <- p0 + 
  geom_line() + 
  theme_minimal()

p2 <- p1 + scale_x_date(
  date_labels = if_else(
    condition = lubridate::month(df$date) == 1,
    true =  "%b\n%Y", 
    false = "%b"
    ),
  date_breaks = "1 months",
  # date_minor_breaks = "1 month",
  expand = c(0,0)
  ) + 
  scale_y_continuous(labels=scales::dollar_format()) +
  xlab(label = "") +
  ylab(label = "Available Funds")

p2
```

```{r}
year_breaks <- df %>%
  slice(2:(n()-1)) %>%         # Remove first and last rows
  filter(month == "January") %>% 
  mutate() %>% 
  pull(date) %>% 
  as.character()
  
p3 <- p2 + geom_vline(
  xintercept = as.Date(year_breaks), 
  linetype="dashed" 
  # color = "red"
  )

p3
```

### Theme
```{r}
theme_custom <- function(){
    ggplot2::theme_classic() +
    ggplot2::theme(
      axis.ticks = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_text(
        hjust = 0
        ),
      axis.text.y = ggplot2::element_text(
        size = 15
        ),
      axis.title.y = ggplot2::element_text(
        size = 20,
        hjust = 0.95,
        vjust = 2
        ),
      plot.margin = unit(c(0, 0.5, 0, 0.5), "cm")
    )
}
```

```{r}
category_key
```


```{r}
p4 <- p3 + theme_custom()
p4
```


```{r}
ggplotly(p4)
```


## Interactive Plot
```{r}
category_filter

df_ia <- projections_raw %>% 
  dplyr::filter(category == category_filter[11])

df_ia
```

```{r}

x <- c(
  "{point.month} {point.year}", 
  "Available Funds"
  )
y <- c(
  "",
  "${point.available_funds:,0f}"
  )


tltip <- highcharter::tooltip_table(x, y)

tltip

highcharter::hchart(
    df_ia,
    "line",
    highcharter::hcaes(
      x = date,
      y = available_funds
    )
) %>%
  highcharter::hc_tooltip(
    useHTML = TRUE,
    headerFormat = "",
    pointFormat = tltip
    ) %>% 
  highcharter::hc_xAxis(
    title = list(
      text = ""
      ),
    labels = list(
      style = list(
        fontSize = "20px"
        )
      )
  ) %>% 
  highcharter::hc_yAxis(
    title = list(
      text = "Available Funds",
      style = list(
        fontSize = "25px"
        )
      ),
    labels = list(
      format = "${value:,0f}",
      style = list(
        fontSize = "20px"
        )
      )
  ) %>% 
   hc_plotOptions(
     line = list(
       lineWidth = 5,
       marker = list(
         lineWidth = 1,
         radius=7.5
       ),
       threshold = 0,
       color = "#0A9396",
       negativeColor = "#9B2226"
     )
   )

```

