---
title: "Exploratory Analysis"
author: "Evan Canfield"
date: "5/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

The purpose of this notebook is to calculate the budget balances on January 1, 2021, and to store it as an internal and external dataset.


# Import
## Packages
```{r import_packages}
# Ensure package installation
if (!require("pacman")) install.packages("pacman")

# Install packages
pacman::p_load(
  dplyr,
  janitor,
  lubridate,
  readxl,
  tidyr
)

```

## Data
```{r}
file_path <- here::here("data-raw", "budgets.xlsx")
budget_default_raw <- readxl::read_xlsx(
  path  = file_path,
  sheet = "default"
)
```


# Internal Datasets
## Category Key
```{r}
# Load and processes the category name look-up table
# Import ----------------------------------------------------------------------
file_path <- here::here("data-raw", "budgets.xlsx")
budget_default_raw <- readxl::read_xlsx(
  path  = file_path,
  sheet = "default"
)

# Processing ------------------------------------------------------------------
## Wide format
category_key <- budget_default_raw %>%
  janitor::clean_names() %>%
  dplyr::select(category) %>%
  dplyr::mutate(
    category_clean = janitor::make_clean_names(category)
  )

# Export-----------------------------------------------------------------------
usethis::use_data(category_key, overwrite = TRUE)
```

## Transactions
### Past
```{r}
# Load and process future transactions data

# Import ----------------------------------------------------------------------
## Transactions

col_type_definition <- list(
  Date =               readr::col_character(),
  Description =        readr::col_character(),
  Amount =             readr::col_double(),
  `Transaction Type` = readr::col_character(),
  Category =           readr::col_character(),
  Note =               readr::col_character()
)

file_path <- here::here("data-raw", "transactions_future.csv")
future_transactions_raw <- readr::read_csv(
  file = file_path, 
  col_types = col_type_definition
  )

# Processing ------------------------------------------------------------------
future_transactions <- future_transactions_raw %>%
  janitor::clean_names() %>%
  dplyr::mutate(
    amount = dplyr::if_else(
      condition = transaction_type == "debit",
      true      = as.double(amount * -1),
      false     = as.double(amount)
    ),
    date = as.Date(date, "%m/%d/%Y"),
    year = lubridate::year(date),
    month = lubridate::month(
      x = date,
      label = TRUE,
      abbr = FALSE
    )
  ) %>%
  dplyr::select(date, amount, category)

# Export-----------------------------------------------------------------------
usethis::use_data(future_transactions, overwrite = TRUE)
```

### Future
```{r}
# Load and process transactions data

# Import ----------------------------------------------------------------------
## Transactions

# Define column types
col_type_definition <- list(
  Date =                   readr::col_character(),
  Description =            readr::col_character(),
  `Original Description` = readr::col_character(),
  Amount =                 readr::col_double(),
  `Transaction Type` =     readr::col_character(),
  Category =               readr::col_character(),
  `Account Name` =         readr::col_character(),
  Labels =                 readr::col_character(),
  Notes =                  readr::col_character()
)

file_path <- here::here("data-raw", "transactions.csv")
transactions_raw <- readr::read_csv(
  file = file_path, col_types = col_type_definition
  )

## Parent / Child Budgets
file_path <- here::here("data-raw", "budgets.xlsx")
budgets_parent_child_raw <- readxl::read_xlsx(
  path  = file_path,
  sheet = "parent_child"
  ) %>%
  janitor::clean_names()


# Processing ------------------------------------------------------------------
## Define override budgets
override_budgets <- c("Travel", "Entertainment")


## Execute processing
transactions <- transactions_raw %>%
  janitor::clean_names() %>%
  dplyr::filter(
    account_name != "CREDITCARD Account"
  ) %>%
  dplyr::mutate(
    amount =  dplyr::if_else(
      condition = transaction_type == "debit",
      true      = as.double(amount * -1),
      false     = as.double(amount)
    ),
    date = as.Date(date, "%m/%d/%Y")
  ) %>%
  dplyr::select(date, amount, category) %>%
  dplyr::left_join(
    budgets_parent_child_raw,
    by = c("category" = "child")
  ) %>%
  dplyr::mutate(
    category =  dplyr::if_else(
      condition = (parent %in% override_budgets),
      true = parent,
      false = category
    )
  ) %>%
  dplyr::select(!parent)


# Export-----------------------------------------------------------------------
usethis::use_data(transactions, overwrite = TRUE)
```

## Budgets
### Budgets - Monthly (Wide Format)
```{r}
# Load and process monthly budget levels, wide format
# Import ----------------------------------------------------------------------
file_path <- here::here("data-raw", "budgets.xlsx")
budget_monthly_wide_raw <- readxl::read_xlsx(
  path  = file_path,
  sheet = "by_month"
)

# Processing ------------------------------------------------------------------
## Wide format
budget_monthly_wide <- budget_monthly_wide_raw %>%
  janitor::clean_names() %>%
  dplyr::select(!c("month", "year")) %>%
  dplyr::select(date, everything())


# Export-----------------------------------------------------------------------
usethis::use_data(budget_monthly_wide, overwrite = TRUE)
```

### Balance on 5/15/2021
```{r}
# Load and process budget balances
# Import ----------------------------------------------------------------------
file_path <- here::here("data-raw", "budgets.xlsx")
budget_balance_raw <- readxl::read_xlsx(
  path  = file_path,
  sheet = "balance"
)

# Processing ------------------------------------------------------------------
budget_balance <- budget_balance_raw %>%
  janitor::clean_names() %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(
      x = date,
      label = TRUE,
      abbr = FALSE
    )
  ) %>%
  dplyr::select(category, balance)


# Export-----------------------------------------------------------------------
usethis::use_data(budget_balance, overwrite = TRUE)
```

### Default Budget Levels
```{r}
# Load and budget defaults data

# Import ----------------------------------------------------------------------
file_path <- here::here("data-raw", "budgets.xlsx")
budget_defaults_raw <- readxl::read_xlsx(
  path  = file_path,
  sheet = "default"
)


# Processing ------------------------------------------------------------------
budget_defaults <- budget_defaults_raw %>%
  janitor::clean_names()


# Export-----------------------------------------------------------------------
usethis::use_data(budget_defaults, overwrite = TRUE)
```

### Baseline Balances
#### Define Cut-off Date
* The cut-off date is May 15, 2021. This was when the Mint budget levels were recorded.
* The start date is the first day of the month of the earliest recorded budget

```{r define_cutoff}
# Define start date
start_date <- budget_monthly_wide %>%
  dplyr::pull(date) %>%
  min() %>%
  as.Date()

# Define cutoff date
cutoff_date <- as.Date("2021-05-15")

# Adjust Cutoff Date by a month
cutoff_date_adj <- cutoff_date - months(1)

print(paste0("Start Date: ", start_date))
```

#### Input Tables
There are three different values we need to derive before we can complete our calculations. 

* Budgeted - Pre Cut-Off: Start date through month before cut-off date
* Budgeted - At Cut-Off: Month of cut-off
* Spending: From start date to cut-off date

##### Amount Budgeted Up to Cut-off Month
We then calculate the amount budgeted in each categor, from the Start Date up to the Cut-off Date. We also exclude the entire month the cut-off date resides in is included in the amount budgeted.
```{r budget_pre_cutoff}
# Extract cleaned names as vectors
categories_clean_vec <- category_key %>% 
  dplyr::pull(category_clean)

# Calculate the total amount budgeted in 2021 up to start of the month of the cut-off date
budgeted_total_by_cutoff <- budget_monthly_wide %>% 
  dplyr::filter(
    # lubridate::year(date)  == lubridate::year(cutoff_date),
    date <= cutoff_date_adj
  ) %>% 
  dplyr::select(dplyr::any_of(categories_clean_vec)) %>% 
  tidyr::pivot_longer(
     dplyr::everything(), 
     names_to = "category_clean", 
     values_to = "budget"
     ) %>% 
  dplyr::group_by(category_clean) %>% 
  dplyr::summarise(budgeted_pre_cutoff = sum(budget)) %>% 
  dplyr::left_join(
    y = category_key, 
    by = "category_clean"
  ) %>% 
  dplyr::select(category, budgeted_pre_cutoff)

budgeted_total_by_cutoff
```

##### Budgets - Month of Cutoff
```{r budget_month_of_cutoff}
budgeted_cutoff_month <- budget_monthly_wide %>% 
  dplyr::filter(
    lubridate::year(date) == lubridate::year(cutoff_date),
    lubridate::month(date) == lubridate::month(cutoff_date)
  ) %>% 
  dplyr::select(!date) %>% 
  tidyr::pivot_longer(
     dplyr::everything(), 
     names_to = "category_clean", 
     values_to = "budgeted_cutoff_month"
  ) %>% 
  dplyr::left_join(
    y = category_key, 
    by = "category_clean"
  ) %>% 
  dplyr::select(category, budgeted_cutoff_month)

budgeted_cutoff_month
```

##### Spending - Up To Cut-off Day
With our spending and budget balance tables properly derived, we will make a table that contains the spending, budgeting, and balance 
```{r spending_cutoff}
# Determine spending on past transactions
category_spend <- transactions %>% 
  dplyr::filter(
    dplyr::between(
      x     = date,
      left  = start_date,
      right = cutoff_date
      )
    ) %>% 
  dplyr::group_by(category) %>% 
  dplyr::summarise(spent = sum(amount))

category_spend
```

#### Compile Budgets and Spending Tables
```{r compile_budget_spending}
# Compile budget balance, spending, and total budgeted
category_summary <- category_spend %>% 
  left_join(
    x = budget_balance, 
    by = "category"
    ) %>% 
  left_join(
    x = budgeted_total_by_cutoff,
    by = "category"
    ) %>% 
  left_join(
    x = budgeted_cutoff_month,
    by = "category"
    ) %>% 
  replace(is.na(.), 0) %>%   # Categories with no spending are NA
  select(category, budgeted_pre_cutoff, budgeted_cutoff_month, spent, balance)

category_summary
```

#### Calculate Available Funds - January 1, 2021
```{r calculate_available_funds}
baseline_funds <- category_summary %>% 
  dplyr::mutate(
    amount = budgeted_cutoff_month - balance - spent -
      (budgeted_pre_cutoff + budgeted_cutoff_month),
    start_date = as.Date(start_date) - days(1)
  ) %>% 
  dplyr::select(start_date, amount, category) %>% 
  dplyr::rename(date = start_date)

baseline_funds
```

#### Save
```{r}
usethis::use_data(baseline_funds,  overwrite = TRUE)
```


# Save to System Data
```{r}
usethis::use_data(
  baseline_funds,
  budget_balance,
  budget_defaults,
  budget_monthly_wide,
  category_key,
  future_transactions,
  transactions,
  internal = TRUE, 
  overwrite = TRUE
  )

```

